\chapter{Introduction}
\label{cha:introduction}

Metric spaces are a generalisation of the world in which we live where
physical objects appear to exist in 3-dimensional space and we have a notion
of the distance between pairs of objects.  The distance that many people think
of is the Euclidean distance, the length of the straight line between the
objects that one would measure with a ruler.  However, taxi drivers are more
familiar with the Manhattan distance, the distance one must travel between
locations by traversing the grid-like streets of a city, and airline pilots
the great circle distance, the shortest distance between two points on a
sphere.  There are many ways to calculate a distance but those most familiar
to humans all share a few key properties, namely they are metrics.  We
therefore live in a metric space, or at least a close approximation of one.

Increasingly, our world is becoming filled with other, more abstract types of
objects that ``live'' outside of the physical space: data.  Regardless of the
form that data takes, be it numbers, text or pictures, the presence of data
always comes with the need for a method of comparison.  Since we are most at
ease with thinking about our own world, it seems natural that we should
imagine data points as ``living'' in a space with a distance defined which
holds the same key properties as the distances we use every day.  Thus, the
data points become members of their own metric space.

There are a number of problems that present themselves when we have a metric
space.  The problem of identifying groups of similar objects based on their
relative proximity is called clustering.  The problem is very old, ubiquitous
and has a rich and diverse set of applications.  Despite the fact that
intelligent beings seem to be naturally adept at the task, it is a well-known
hard problem for a computer, and in more than one sense.  The first difficulty
is in defining precisely what one actually expects in terms of a metric,
rather than the vague objective of ``groups of similar objects''.  The second
difficulty is in the sense of computational complexity.  In the first part of
this thesis we focus mainly on the first difficulty in the context of
partitional clustering, which seeks to partition a set into a given number of
subsets.  Many objective criteria have been developed to measure the
usefulness of a partition as a solution to a given clustering problem, but we
show that even seemingly similar optimisation criteria can produce vastly
different results (see Section~\ref{sec:worst-case-perf}).

In order to show that criteria can produce vastly different partitions it is
necessary to be able to compare partitions.  Some metrics have been devised
for that purpose already but they are all naïve with respect to the fact that
our data itself lives in a metric space.  We find ourselves with several
layers of metric spaces and we introduce the concept of ``lifting'' a metric
space to the space of its power set.  In this way we introduce a new metric
for comparing partitions that can take into account the fact that the data
lives in a metric space (see Section~\ref{sec:lifting-metric-space}).

Metrics present themselves in the context of that most well-known and useful
data structure: the tree.  In a rooted edge-weighted $X$-tree, that is a tree
with leaf set $X$, the graph-theoretic distance between the leaves is a metric
called a tree metric.  It is well known that a tree is uniquely defined by and
can be reconstructed from the tree metric induced on its leafset.  Tree
construction is appropriate and desirable in many areas of classification such
as in evolutionary biology.  For example, if we had a dataset corresponding to
varieties of some organism, we could find a partitional clustering on them in
which each cluster corresponds to a continent, reflecting the expected result
that organisms on the same continent are more closely related, and those on
different continents are more distantly related.  However, a tree can show us
not only this information but the relationships and lineages of each variety
in our set.  In this way, trees can be considered hierarchical clusterings and
a step up from partitional clusterings.

In practice, though, we may not always have access to the complete distance
information for a set of elements.  Rather we might be presented with a
partial distance, that is where the distance between some pairs elements is
unknown.  Studying the properties of partial distances and their ability to
uniquely identify a tree has given rise to the theory of ``lassoing'' a tree.
It turns out that certain subsets of all pairwise distances do indeed contain
enough information to uniquely identify either the topology of a tree, its
edge weights, or both.  This theory is reviewed in
Section~\ref{sec:lassoing-corralling}.

We therefore investigate the problem of constructing an edge-weighted tree
from partial distance information.  We begin by investigating the properties
of a special type of lasso and then we present an algorithm for constructing
the unique tree which corresponds to such a lasso.  We then turn to the more
general problem of reconstruction from partial distance information and
present an algorithm which constructs a tree from a partial distance and
returns the set of given distances which uniquely determine the constructed
tree.

This thesis is in two parts with the first part focusing on partitional
clustering and the second on reconstruction of hierarchical clusterings
(trees) from partial distance information.  It is organised as follows: in
Chapter~\ref{cha:background}, we introduce the concept of metric spaces and
review various metrics that have been used over the years.  We then introduce
partitions and review metrics for comparing partitions.  This leads to a
review of partitional clustering, the problem of finding partitions of a
dataset.  Finally, we introduce trees and their applications.  In
Chapter~\ref{cha:sum-squar-clust}, we focus on partitional clustering and its
difficulties by comparing and contrasting two closely related partitional
clustering criteria, called sum-of-squares criteria.  We show that the two
criteria can produce completely different results.  In
Chapter~\ref{cha:background2}, we turn our attention to the tree
reconstruction problem and introduce trees, tree metrics, tree reconstruction
and lassos.
% In Chapter~\ref{cha:dist-minim-topl}, we investigate the
% properties of a special kind of lasso which we call a distinguished minimal
% topological lasso and present an algorithm for tree construction from such a
% lasso.
In Chapter~\ref{cha:lasso-construction}, we present a general algorithm called
\textsc{Lasso} for tree reconstruction from partial distances which is
consistent according to the theory of lassos and show that it is applicable to
construction of trees and supertrees from partial distance information.  We
apply Finally we conclude the thesis and suggest further work in
Chapter~\ref{cha:conclusion}.

%%% Local Variables:
%%% TeX-master: "thesis"
%%% End:
